---
title: "Milestone 1"
author: "Almas K."
date: '2020-02-29'
output:
  html_document:
    keep_md: yes
---

# Alcohol Use and Association with Final Grades 

```{r}
library(tidyverse)
library(here)
library(knitr)
library(corrplot)
library(patchwork)
```

## Introduction:

 Alcohol use has been linked with cognitive impairement in the short term in a variety of situations such as in operation of a motor vehicle.Its effect on academic tasks would be an interesting area to explore.  Already, numerous factors have been found to affect a student's performance in a class, from sleep to diet. It would be interesting to see how alcohol use has an effect on the cognitive performance of students over a longer period of time, such as in the form of a final grade or overall performance in a class. 
## Data Description
The datasets are obtained from UCI  and is originally from Fabio Pagnotta and Hossain Mohammad Amran. It contains survey data from Portugese highschool students in a Math and Portugese class and contains information on 33 attributes. Each class is its own .csv file, but I will be focussing on the attributes from the Portugese class dataset as it contains more students (649 students). Each student makes up each row. Below is the entire variable set:
This was generated from a colon separated file I made from the original txt metadata file.

```{r}
meta_dat <- read.delim((here("data","student_metadata.txt")),sep = ";", header=FALSE)
colnames(meta_dat) <- c("variable","description","type")
knitr:::kable(meta_dat)
```



## Exploring the Dataset

### Load data

```{r}
library(here)
stu_por <-read.csv(here("data","student-por.csv"),sep=";")
str(stu_por)
```

```{r}
any(is.na(stu_por)) # checks for missing values
```

### Correllogram
```{r}
student_cor <- stu_por%>%
  select_if(is.numeric) %>% # select numeric values
  sapply(as.double) %>% # convert to double
  cor() # run correlation

student_cor<- round(student_cor,2)
corr <- corrplot(student_cor, 
         type="upper", 
         method="color",
         tl.srt=45, 
         tl.col = "blue",
         diag = FALSE) 
```

G1 and G2 having the highest correlation with G3 makes sense here, as earlier term grades are correlated with later term grades.





### Boxplots
Let's look at weekend alcohol and workday alcohol use's spread.

```{r}

walcstu_box <- stu_por %>%
  ggplot(aes(y=G3,group=Walc,fill=as.factor(Walc))) +
  geom_boxplot()+ 
  ggtitle("Final grade and Weekend Alcohol Consumption")

dalcstu_box <- stu_por %>%
  ggplot(aes(y=G3,group=Dalc,fill=as.factor(Dalc))) +
  geom_boxplot()+
   ggtitle("Final grade and Workday Alcohol Consumption")

dalcstu_box 
walcstu_box
   
```


### Density Plots
Let's look at the distribution of grades.
```{r}
 # Let's look at the distribution of grades
    stu_por %>%
  ggplot(aes(x=G3)) +
  geom_density()+
  ggtitle("Spread of Final Grades")
```

The distribution of grades appear to be a bit left skewed. 

#### Other variables that may affect data
Let's look at potential confounding factors like sex of the student, parental status and family support and their spread in average final grades
```{r}
Sex_grades <- stu_por %>%
  ggplot(aes(y=G3,col=sex)) +
  geom_boxplot()+
  ggtitle("Sex of Individual")

Pstat_grades <-  stu_por %>%
  ggplot(aes(y=G3,col=Pstatus)) +
  geom_boxplot()+
  ggtitle('Parental Status')

Famsup_grades <-  stu_por %>%
  ggplot(aes(y=G3,col=as.factor(famsup))) +
  geom_boxplot()+
  ggtitle('Family Support')

health_stat_grades <-  stu_por %>%
  ggplot(aes(y=G3,col=as.factor(health))) +
  geom_boxplot()+
  ggtitle('Health')

(Sex_grades+Pstat_grades)/(Famsup_grades+health_stat_grades) +
 plot_annotation(title="Potential Confounding factors and grades")
```

It doesn't look like there is a huge difference between the grades in males compared to females. Males have a slighly lower average, but overall are similar. This is good because it will not be a huge confound in the data.Also family support and parental status have similar average values. 

## Research Question:

In this analysis, I will use linear regression to determine the relationship between alcohol use, either weekend, weekday (workday) or both and final grades (G3) for students. I chose the final grades as a output variable because it is more resistant to short term effects because it depends on work throughout the term. 


## Plan of Action:

I will remove those with very bad health status (1), as to reduce confounds in the data. My main focus is on the alcohol use categories and final grades, so I will probably ignore the other factors.I will then perform linear regression analysis and plot 
a regression line using the relevant variables. 


